---
title: "Exploratory data analysis for 554 Survey"
output: github_document
---

```{r import}
suppressPackageStartupMessages(library(tidyverse))
```

## Load Data

```{r Load Data}
# load csv: note the data is not in the github repo
df = read.csv("results-survey.csv")
# select the columns we need
df <- df %>% select(id, X1, X2, X3, X4, Y)
# display a head
head(df)
```

## Variables Collected

The data is collected from the response of our survey. Here is the decriptions of the variables.

**X1:** What is the highest degree you obtained before MDS?

- A1: Bachelor's
- A2: Master's or Higher

**X2:** What was your most recent degree in?

- A1: Computer Science
- A2: Engineering (Other than Computer Science)
- A3: Statistics
- A4: Mathematics (Other than Statistics)
- A5: None of the above

**X3:** How related was you work experience to the MDS program?

- A1: Very related - the work I did was very similar to the work I am doing in the program
- A2: Somewhat related - the work I did was somewhat similar to the work I am doing in the program
- A3: Not related -  the work I did was not at all similar to the work I am doing in the program
- A4: I have no previous work experience

**X4:** Did you have programming experience before this program?

- A1: Yes
- A2: No

**Y:** How difficult have you found the MDS program?

- 1: Very Easy
- 2: Easy
- 3: Neutral
- 4: Difficult
- 5: Very Difficult

## Summary table

```{r}
# summarize the data
summary(df)
```

## Bar plots for the single variables

```{r X1}
df %>% 
  ggplot(aes(x = X1)) +
  geom_bar(width = 0.4, fill = "coral2") +
  scale_x_discrete(labels = c("Bachelor's","Master's or Higher")) +
  labs(title = "What is the highest degree you obtained before MDS?",
       x = "Degree",
       y = "Count")
```

```{r X2}
df %>% 
  ggplot(aes(x = X2)) +
  geom_bar(width = 0.6, fill = "cadetblue3") +
  scale_x_discrete(labels = c("CS","Engineering","Stats","Math","Others")) +
  labs(title = "What was your most recent degree in?",
       x = "Major",
       y = "Count")
```

```{r X3}
df %>% 
  ggplot(aes(x = X3)) +
  geom_bar(width = 0.6, fill = "darkseagreen3") +
  scale_x_discrete(labels = c("Very related","Somewhat related","Not related","No work experience")) +
  labs(title = "How related was you work experience to the MDS program?",
       x = "Relationship",
       y = "Count")
```

```{r X4}
df %>% 
  ggplot(aes(x = X4)) +
  geom_bar(width = 0.4, fill = "goldenrod2") +
  scale_x_discrete(labels = c("Yes","No")) +
  labs(title = "Did you have programming experience before this program?",
       x = "Experience",
       y = "Count")
```

```{r Y}
df %>% 
  ggplot(aes(x = Y)) +
  geom_bar(width = 0.6, fill = "slateblue") +
  labs(title = "How difficult have you found the MDS program? (1-5)",
       x = "Difficulty",
       y = "Count")
```